---
title: "National Climatic Data Center Storm Events Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data 

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

## Environment Setup

```{r Library and Setup, message=FALSE}
library(tidyverse)
library(tidytext)
library(dygraphs)
library(lubridate)
library(stringi)
sessionInfo()
```

## Load Data

```{r Load Data, message=FALSE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fname <- c("./data/maindf.csv.bz2")

if(!file.exists("data")){
        dir.create("data")
}

download.file(fileUrl,destfile="./data/maindf.csv.bz2",method="curl")

maindf <- read.csv(fname, header = TRUE, na.strings=c("","NA"))
dateDownloaded <- date()
dateDownloaded
```

Maintain the original "maindf" data frame.  

```{r, cache=TRUE}
df <- maindf

head(df)

# sum(is.na(df$END_TIME))
# sum(is.na(df$END_DATE))
# range(mdy(df$BGN_DATE)
```

## Purpose

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

Data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

### Strategy in Data Wrangling

1. Analyze missing data
2. Analyze Date and time
3. Analyze EVTYPE



```{r Missing_Data-Analysis}
naAnalysis <- df %>%
    purrr::map_df(function(x) round(mean(is.na(x)),digits = 2)*100) %>%
    gather(EVType, naAverage) %>% 
    print(n = 10)

naAnalysis1 <- naAnalysis %>% filter(naAverage>0) 
quantile(naAnalysis1$naAverage) # range 16 100
```


```{r Missing_Data_plot}
naAnalysis1 %>% ggplot(aes(x=EVType, y=naAverage)) + geom_point() + theme(axis.text.x=element_text(angle=90,hjust=.1))+labs(x= "Event Type", y= "NA Average (%)", title = "Missing Data Analysis")
```
We can see that out of the original 37 variables, 14 variables has missing data. 







